\begin{acknowledgements}\label{acknow}
\ssp

\noindent First and foremost, I would like to acknowledge the support and guidance of my \\ coadvisors, Drs. Dan Gillis and Bob Hanner for their encouragement over the years. In addition, thanks go out to all current and past members of the Gillis and Hanner Lab groups, particularly Rob Young for all the fun times that were had throughout the journey.

\vspace{5mm}

\noindent Secondly, I wish to thank Dr. Deb Stacey and Dr. Graham Taylor for serving on my advisory committee, graciously reading my individual manuscripts and the current thesis, as well as engaging in stimulating discussion during my many committee meetings and graduate seminars over the past four and a half or so years.

\vspace{5mm}

\noindent Third, I would like to recognize the tremendous efforts of both Steven French and Scarlett Bootsma for their interest and willingness to learn about the inner workings of the {\tt HACSim} algorithm and assist with coding, debugging and simulation as part of their fourth year thesis projects.

\vspace{5mm}

\noindent Fourth, thanks to Sally Adamowicz and Rodger Gwiazdowski for providing \\ valuable edits and comments to individual chapters, published or otherwise, found within this thesis.

\vspace{5mm}

\noindent Fifth, a big shout out goes out to graduate students in the Adamowicz, Hebert \\ Hajibabaei and Steinke Labs for their friendships throughout the years, especially Jacopo D'Ercole and Mike Wright.

\vspace{5mm}

\noindent Thanks also go out to Jennifer Hughes, for addressing my many inquiries over the years regarding administrative aspecies of Computational Sciences PhD. Program.

\vspace{5mm}

\noindent I sincerely apologize in advance to those whom I may have missed in acknowledging here. Just know that whatever your contribution, no matter how big or small, to my success in this endeavour, Thank You!

\vspace{5mm}

\noindent Lastly, I would like to thank my family for their love and encouragement throughout my journey into the vast unknown that is doctoral study. Now they can finally stop constantly asking: ``When are you going to get your PhD.?!"

\vspace{5mm}

\noindent This work was supported in part by a Graduate Excellence Entrance (GEE) \\ Scholarship. Other sources of funding came from various travel grants which greatly aided the ability to attend and present my research at the 7th and 8th International Barcode of Life (iBOL) Conferences held in Kruger National Park, South Africa and Trondheim, Norway in 2017 and 2019 respectively.


\end{acknowledgements}
\dsp
